Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit

### 2023-07-22 Manuscript/Manuscript.tex
Ms: knit

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  10pt,
  onecolumn]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Arial}
  \setmonofont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=0.8in,right=0.8in,top=0.8in,bottom=0.5in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Nitric oxide (NO) produced by nitric-oxide synthase (NOS) is a key
regulator of animal physiology. Here we uncover a function for NO in the
integration of UV exposure and the gating of a UV-avoidance circuit. We
studied UV/violet avoidance mediated by brain ciliary photoreceptors
(cPRCs) in larvae of the annelid \emph{Platynereis dumerilii}. In the
larva, NOS is expressed in interneurons (INNOS) postsynaptic to cPRCs.
UV stimulation of cPRCs triggers INNOS activation and NO production. NO
signals retrogradely to cPRCs to induce their sustained post-stimulus
activation through an unconventional guanylyl cyclase. This late
activation inhibits serotonergic ciliomotor neurons to induce downward
swimming. In \emph{NOS} mutants, retrograde signalling, circuit output
and UV avoidance are defective. By mathematical modelling, we
recapitulate phototransduction and circuit dynamics in wild-type and
mutant larvae. Our results reveal how NO-mediated retrograde signalling
gates a synaptic circuit and induces short-term memory of UV exposure to
orchestrate light-avoidance behaviour.

@@

Nitric oxide (NO) is one such modulator with unique physical and
signalling properties. This free radical synthesized from L-arginine by
nitric oxide synthase (NOS) is short-lived and can diffuse across
biological membranes (Cudeiro and Rivadulla, 1999; Thomas, 2015).
Canonical NO signalling involves the Ca2+/calmodulin-dependent
activation of NOS, NO production and diffusion, and the NO-dependent
activation of soluble guanylate cyclases (sGC). This can occur
cell-autonomously or in other cells leading to cGMP production (Bredt et
al., 1990; Hölscher, 1997). Given that NOS activation is calcium
dependent and NOS shows neuron-type-specific expression (Aso et al.,
2019; Gibbs and Truman, 1998; Mobley et al., 2022; Wildemann and Bicker,
1999), NO action can lead to the activity-dependent modulation of neural
circuits at specific sites (Aso et al., 2019; Jacoby et al., 2018;
Vielma et al., 2014; Wang et al., 2007) .

In the vertebrate retina, NOS is expressed in amacrine, ganglion and
other cells and the actions of NO can be diverse (Cudeiro and Rivadulla,
1999; Jacoby et al., 2018; Wang et al., 2007). For example, defective NO
signalling in NOS knockouts leads to a decreased sensitivity of retinal
ganglion cells to light stimulation (Wang et al., 2007). In the retina,
NO signalling can also involve pathways other than canonical sGC
signalling (Jacoby et al., 2018; Tooker et al., 2013; Wei et al., 2012).
Due to the complex expression of NOS in vertebrates and the diversity of
its functions, it has been challenging to link the neurophysiological
effects of NO to signalling mechanisms and behaviour change.

@@

Here we uncovered an essential function for NO signalling in larval
UV/violet-light avoidance behaviour. In \emph{Platynereis} larvae,
UV/violet avoidance is mediated by brain ciliary photoreceptor cells
(cPRCs) and is characterised by downward swimming (Verasztó et al.,
2018). The cPRCs express a ciliary-type opsin, c-opsin1 (Arendt et al.,
2004) that forms a UV-absorbing bistable photopigment with an absorption
maximum around 384 nm (Tsukamoto et al., 2017; Veedin Rajan et al.,
2021; Verasztó et al., 2018). Upon UV/violet exposure, the cPRC show a
characteristic biphasic calcium response that is \emph{c-opsin1}
dependent. UV/violet avoidance is also \emph{c-opsin1}-dependent and is
defective in \emph{c-opsin1} mutants (Verasztó et al., 2018). Here we
show that \emph{Platynereis} \emph{NOS} is expressed in interneurons of
the cPRC circuit and is required for UV/violet-avoidance. By combining
calcium imaging across the fully-mapped cPRC circuit (Verasztó et al.,
2018) with genetic perturbations and mathematical modelling, we describe
how NO tunes circuit dynamics through non-synaptic retrograde signalling
to cPRCs. This delayed neuroendocrine feedback integrates UV/violet
exposure to induce a short-term memory manifested in altered circuit
activity and an aversive behavioural response.

@@

We identified a single \emph{nitric oxide synthase} (\emph{NOS}) gene in
the \emph{Platynereis dumerilii} genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that \emph{Platynereis}
NOS belongs to an orthology group of bilaterian NOS sequences (Figure
1---figure supplement 1). To characterise the expression pattern of
\emph{NOS} we used in situ hybridization chain reaction (HCR) and
transient transgenesis. In two- and three-day-old larvae, we detected
\emph{NOS} expression in four cells (two of them weakly expressing) in
the apical organ region (Figure 1D and Figure 1---figure supplement 2).
\emph{NOS} was also expressed in the region of the visual eyes (adult
eyes) and the pigmented eyespots (Figure 1---figure supplement 2). The
four apical organ cells, but not the eyes, were also labelled with a
\emph{NOS} reporter construct driving palmitoylated-tdTomato (Figure
1E). This reporter also revealed the axonal projections of these central
NOS-expressing neurons. The position and morphology of the four
\emph{NOS+} cells allowed us to identify the same four cells as four
interneurons (INNOS) in our three-day-old whole-body \emph{Platynereis}
volume EM data (Verasztó et al., 2020; Williams et al., 2017) (Figure
1B,C). In the synaptic connectome, the INNOS cells are postsynaptic to
the UV-sensory cPRCs and presynaptic to the INRGWa interneurons, which
are also cPRC targets (Figure 1C,F). The projections of INNOS cells are
segregated into input (dendritic) and output (axonal) compartments, with
cPRC inputs occurring on the dendritic part in the dense neurosecretory
plexus. INNOS output synapses form in the more ventral projection region
(Figure 1---figure supplement 3). INRGWa neurons synapse on the head
serotonergic ciliomotor neurons (Ser-h1), which synapse on the
prototroch ciliary band and the cholinergic MC ciliomotor neurons
(Figure 1C,F) (Verasztó et al., 2017).

\begin{figure}
\includegraphics[width=25in]{figures/Fig1} \caption{**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** Scanning electron microscopy image of a three-day-old *Platynereis* larva. **(B, C)** Volume rendering of the neuron types (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed from a whole-body transmission electron microscopy volume of a three-day-old larva. Neurite skeletons are shown with cell-body positions represented by spheres. Projections of all neurons in the body are shown in grey to highlight the neuropils. The outline of the yolk is also indicated in grey. In B, nuclei positions of the prototroch head ciliary band are shown as grey spheres. **(D)** Expression of the *NOS* gene detected by in situ HCR (magenta) in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(E)** Expression of a *NOS* reporter (NOSp::palmi-3xHA: magenta) labelled with an anti-HA antibody in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(F)** Synaptic wiring diagram of the cPRC circuit. Hexagons represent cell groups, with the number of cells per group shown in square brackets. Arrows represent the summed number of synaptic contacts between cell groups. Arrow thickness is proportional to the number of synapses.}\label{fig:unnamed-chunk-1}
\end{figure}

@@

The expression of \emph{NOS} in the INNOS interneurons in the cPRC
circuit suggests that NO signalling may be involved in
UV/violet-avoidance. To test this, first we asked whether NO is produced
during UV/violet stimulation of the larvae. We injected the fluorescent
NO-reporter DAF-FM into zygotes and imaged two-day-old larvae while
exposing the region of cPRC cilia to 405 nm violet light. To mark cell
outlines, we coinjected mRNA encoding a red-fluorescent reporter
(RGECO), allowing the identification of the cPRCs. Following light
stimulation in the region of the ramified cilia of the cPRCs, we
detected an increase in DAF-FM fluorescence in the anterior
neurosecretory neuropil, the region of INNOS projections. This increase
did not occur in larvae where we illuminated a control area (Figure 2).

\begin{figure}
\includegraphics[width=29.17in]{figures/Fig2} \caption{**Figure 2. NO produced by UV/violet stimulation to cPRCs.** **(A)** DAF-FM fluorescence in the region of the neurosecretory neuropil. The white line indicates the outline of the larva. The dashed line corresponds to the area where fluorescence was quantified. Circles indicate the location of cPRC and control stimulation. cPRCs are marked by thin lines. **(B)** Changes in DAF-FM fluorescence before and during 405 nm light stimulation. **(C)** Changes in DAF-FM fluorescence over time during 405 nm stimulation of the cPRCs (green) or a control area (ctr: gray). The purple box indicates the duration of 405 nm stimulation. Individual traces normalized (ΔF/F0) are shown as thin lines. Thick lines show the mean value with 0.95 confidence intervals. N = 9 larvae for control and 11 for cPRC stimulation.}\label{fig:unnamed-chunk-2}
\end{figure}

@@

We next tested whether NO signalling regulates UV/violet avoidance. To
achieve this, we generated two \emph{Platynereis} \emph{NOS} knockout
lines with the CRISPR/Cas9 method. We recovered two deletions
(NOSΔ11/Δ11 and Δ23/Δ23), both frame-shift mutations leading to an early
stop codon and thus likely representing null alleles (Figure 3---figure
supplement 1A). We could establish a homozygous line for both mutations
indicating that NOS is not an essential gene in \emph{Platynereis}. To
quantify UV avoidance, we recorded the trajectories of freely swimming
wild type and mutant larvae in vertical columns, illuminated laterally
from two opposite sides with 395 nm UV light (Figure 3A and Figure
3---figure supplement 1B). As previously shown, wild-type larvae swim
downward following non-directional UV/violet light stimulation (Verasztó
et al., 2018). In contrast, both two- and three-day-old homozygous
\emph{NOS}-mutant larvae showed a strongly diminished UV-avoidance
response (Figure 3A, B and Figure 3---figure supplement 1B,C). This
phenotype is similar to the defective UV-avoidance of \emph{c-opsin1}
mutant larvae (Verasztó et al., 2018) and reveals a requirement for NOS
in UV-avoidance behaviour. Wild type but not mutant larvae also showed
an increase in swimming speed under UV light that may be due to downward
swimming trajectories (swimming in the direction of gravity) (Figure 3B
and Figure 3---figure supplement 1C). We also tested directional
phototaxis, by exposing larvae to 480 nm directional collimated light
from the top of the column. Three-day-old but not two-day-old
\emph{NOS}-mutant larvae also showed reduced phototactic behaviour,
suggesting a function for \emph{NOS} in the visual eyes that mediate
three-day-old phototaxis (Randel et al., 2014) (Figure 3D and Figure
3---figure supplement 1G).

To distinguish between an acute and developmental function of NOS in
light responses, we next tested larvae exposed to the NOS inhibitor
L-NAME. Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a
dose-dependent inhibition of UV avoidance. In contrast, phototaxis was
not affected (Figure 3C, E). Overall, our results indicate an acute
requirement for NOS signalling in UV-avoidance and a possible indirect,
developmental role in the visual system, reminiscent of the function of
NO signalling in \emph{Drosophila} eye development (Gibbs and Truman,
1998).

\begin{figure}
\includegraphics[width=33.33in]{figures/Fig3} \caption{**Figure 3. NOS is required for UV avoidance in *Platynereis* larvae.**  **(B)** Swimming trajectories of wild type (WT, n=32) and NOS mutant (NOSΔ11/Δ11, n=26 and NOSΔ23/Δ23, n=47) three-day-old larvae. All trajectories start at 0 x and y position and time 0 corresponding to 10 sec after the onset of 395 nm stimulation from the side. **(B)** Vertical position of batches of wild type and mutant three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. **(C)** Vertical position of batches of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. **(D)** Vertical displacement in 30 sec bins of wild type and mutant (NOSΔ11 and NOSΔ23) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. **(E)** Vertical displacement in 30 sec bins of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top.}\label{fig:unnamed-chunk-3}
\end{figure}

@@

To investigate how NO signalling alters the dynamics of the cPRC
circuit, we carried out calcium imaging experiments. We ubiquitously
expressed the calcium sensor GCaMP6s in larvae and imaged calcium
signals during 405 nm stimulation of the cPRCs. As we have shown
previously, a 20-sec local stimulation of cPRC cilia led to a transient
increase in cPRC calcium levels, followed by a transient decrease
(Verasztó et al., 2018). After \textasciitilde20-sec, calcium levels in
cPRCs were raising again, reaching higher levels than at the start of
the stimulus -- a response that may involve depolarisation (Figure 4A,
B). This activation phase occurs after the 20 sec stimulation period and
is likely due to a delayed neuroendocrine feedback (Verasztó et al.,
2018). To determine whether NO mediates such a feedback, we repeated the
experiment in \emph{NOS}-mutant larvae. While we detected the initial
activation phase followed by inhibition, in homozygous
\emph{NOS}-mutants for both CRISPR alleles this was not followed by
delayed activation. Instead, calcium levels dropped to a low
steady-state level (Figure 4A, B). We thus identified a requirement for
NO signalling in the late-phase activation of cPRCs.

\textbf{Two unconventional guanylyl cyclases are expressed in the cPRCs}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  10pt,
  onecolumn]{article}
\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{Arial}
  \setmonofont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=0.8in,right=0.8in,top=0.8in,bottom=0.5in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\usepackage{svg}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Nitric oxide (NO) produced by nitric-oxide synthase (NOS) is a key
regulator of animal physiology. Here we uncover a function for NO in the
integration of UV exposure and the gating of a UV-avoidance circuit. We
studied UV/violet avoidance mediated by brain ciliary photoreceptors
(cPRCs) in larvae of the annelid \emph{Platynereis dumerilii}. In the
larva, NOS is expressed in interneurons (INNOS) postsynaptic to cPRCs.
UV stimulation of cPRCs triggers INNOS activation and NO production. NO
signals retrogradely to cPRCs to induce their sustained post-stimulus
activation through an unconventional guanylate cyclase. This late
activation inhibits serotonergic ciliomotor neurons to induce downward
swimming. In \emph{NOS} mutants, retrograde signalling, circuit output
and UV avoidance are defective. By mathematical modelling, we
recapitulate phototransduction and circuit dynamics in wild-type and
mutant larvae. Our results reveal how NO-mediated retrograde signalling
gates a synaptic circuit and induces short-term memory of UV exposure to
orchestrate light-avoidance behaviour.

@@

Nitric oxide (NO) is one such modulator with unique physical and
signalling properties. This free radical synthesized from L-arginine by
nitric oxide synthase (NOS) is short-lived and can diffuse across
biological membranes (Cudeiro and Rivadulla, 1999; Thomas, 2015).
Canonical NO signalling involves the
Ca\textsuperscript{2+}/calmodulin-dependent activation of NOS, NO
production and diffusion, and the NO-dependent activation of soluble
guanylate cyclases (sGC) leading to cGMP production (Bredt et al., 1990;
Hölscher, 1997). Given that NOS activation is Ca\textsuperscript{2+}
dependent and NOS shows neuron-type-specific expression (Aso et al.,
2019; Gibbs and Truman, 1998; Mobley et al., 2022; Wildemann and Bicker,
1999), NO action can lead to the activity-dependent modulation of neural
circuits at specific sites (Aso et al., 2019; Jacoby et al., 2018;
Vielma et al., 2014; Wang et al., 2007) .

In the vertebrate retina, NOS is expressed in amacrine, ganglion and
other cells and the actions of NO can be diverse (Cudeiro and Rivadulla,
1999; Jacoby et al., 2018; Wang et al., 2007). For example, defective NO
signalling in NOS knockout mice leads to a decreased sensitivity of
retinal ganglion cells to light stimulation (Wang et al., 2007). In the
retina, NO signalling can also involve pathways other than canonical sGC
signalling (Jacoby et al., 2018; Tooker et al., 2013; Wei et al., 2012).
Due to the complex expression of NOS in vertebrates and the diversity of
its functions, it has been challenging to link the neurophysiological
effects of NO to signalling mechanisms and behaviour change.

@@

We uncovered an essential function for NO signalling in larval
UV/violet-light avoidance behaviour. In \emph{Platynereis} larvae,
UV/violet avoidance is mediated by brain ciliary photoreceptor cells
(cPRCs) and is characterised by downward swimming (Verasztó et al.,
2018). The cPRCs express a ciliary-type opsin, c-opsin1 (Arendt et al.,
2004) that forms a UV-absorbing bistable photopigment with an absorption
maximum around 384 nm (Tsukamoto et al., 2017; Veedin Rajan et al.,
2021; Verasztó et al., 2018). Upon UV/violet exposure, the cPRCs show a
characteristic biphasic Ca\textsuperscript{2+} response that is c-opsin1
dependent. UV/violet avoidance is also c-opsin1 dependent and is
defective in \emph{c-opsin1} mutants (Verasztó et al., 2018). Here we
show that \emph{Platynereis} NOS is expressed in interneurons of the
cPRC circuit and is required for UV/violet-avoidance. By combining
Ca\textsuperscript{2+} imaging across the fully-mapped cPRC circuit
(Verasztó et al., 2018) with genetic perturbations and mathematical
modelling, we describe how NO tunes circuit dynamics through
non-synaptic retrograde signalling to cPRCs. This delayed neuroendocrine
feedback integrates UV/violet exposure to induce a short-term memory
manifested in altered circuit activity and an aversive behavioural
response.

@@

We identified a single \emph{nitric oxide synthase} (\emph{NOS}) gene in
the \emph{Platynereis dumerilii} genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that \emph{Platynereis}
NOS belongs to an orthology group of bilaterian NOS sequences (Figure
1---figure supplement 1). To characterise the expression pattern of
\emph{NOS}, we used in situ hybridization chain reaction (HCR) and
transient transgenesis. In two- and three-day-old larvae, we detected
\emph{NOS} expression in four cells (two of them weakly expressing) in
the apical organ region (Figure 1D and Figure 1---figure supplement 2).
\emph{NOS} was also expressed in the region of the visual eyes (adult
eyes) and the pigmented eyespots (Figure 1---figure supplement 2). The
four apical organ cells, but not the eyes, were also labelled with a
\emph{NOS} reporter construct driving palmitoylated tdTomato (Figure
1E). This reporter also revealed the axonal projections of these central
NOS-expressing neurons. The position and morphology of the four
\emph{NOS+} cells allowed us to identify the same four cells as four
interneurons (INNOS) in our three-day-old whole-body \emph{Platynereis}
volume EM data (Verasztó et al., 2020; Williams et al., 2017) (Figure
1B,C). In the synaptic connectome, the INNOS cells are postsynaptic to
the UV-sensory cPRCs and presynaptic to the INRGWa interneurons, which
are also cPRC targets (Figure 1C, F). The projections of INNOS cells are
segregated into input (dendritic) and output (axonal) compartments, with
cPRC inputs occurring on the dendritic part in the dense neurosecretory
plexus. INNOS output-synapses form in the more ventral projection region
(Figure 1---figure supplement 3). INRGWa neurons synapse on the head
serotonergic ciliomotor neurons (Ser-h1), which synapse on the
cholinergic MC ciliomotor neuron and cells of the prototroch ciliary
band (Figure 1C, F) (Verasztó et al., 2017).

\begin{figure}
\includegraphics[width=25in]{figures/Fig1} \caption{**Figure 1. Identification of *NOS*-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** Scanning electron microscopy image of a three-day-old *Platynereis* larva. **(B, C)** Volume rendering of the neuron types (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed from a whole-body transmission electron microscopy volume of a three-day-old larva. Neurite skeletons are shown with cell-body positions represented by spheres. Projections of all neurons in the body are shown in grey to highlight the neuropils. The outline of the yolk is also indicated in grey. In B, nuclei positions of the prototroch head ciliary band are shown as grey spheres. **(D)** Expression of the *NOS* gene detected by in situ HCR (magenta) in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(E)** Expression of a *NOS* reporter (NOSp::palmi-3xHA; magenta) labelled with an anti-HA antibody in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub; green) highlights cPRC cilia and the neuropil. **(F)** Synaptic wiring diagram of the cPRC circuit. Hexagons represent cell groups, with the number of cells per group shown in square brackets. Arrows represent the summed number of synaptic contacts between cell groups. Arrow thickness is proportional to the number of synapses.}\label{fig:unnamed-chunk-1}
\end{figure}

@@

The expression of \emph{NOS} in the INNOS interneurons in the cPRC
circuit suggests that NO signalling may be involved in
UV/violet-avoidance. To test this, first we asked whether NO is produced
during UV/violet stimulation of the larvae. We injected the fluorescent
NO-reporter DAF-FM into zygotes and imaged two-day-old larvae while
exposing the region of cPRC cilia to 405 nm violet light. To mark cell
outlines, we coinjected mRNA encoding a red-fluorescent reporter
(RGECO), allowing the identification of the cPRCs. Following light
stimulation in the region of the ramified cilia of the cPRCs, we
detected an increase in DAF-FM fluorescence in the anterior
neurosecretory neuropil, the region of INNOS projections. This increase
did not occur in larvae in which we illuminated a control area (Figure
2).

\begin{figure}
\includegraphics[width=29.17in]{figures/Fig2} \caption{**Figure 2. NO produced by UV/violet stimulation to cPRCs.** (**A**) DAF-FM fluorescence in the region of the neurosecretory neuropil. The white line indicates the outline of the larva. The dashed line corresponds to the area where fluorescence was quantified. The circles indicate the location of cPRC and control stimulation. The cPRCs are marked by thin lines. (**B**) DAF-FM fluorescence before and during 405 nm light stimulation. (**C**) Changes in DAF-FM fluorescence over time during 405 nm stimulation of the cPRCs or a control area (ctr stim.). The purple box indicates the duration of 405 nm stimulation. Individual normalized traces (ΔF/F0) are shown as thin lines. Thick lines show the mean value with 0.95 confidence intervals. N = 9 larvae for control and 11 for cPRC stimulation. <br> Figure 2 -- source data 1. DAF-FM fluorescence reads.}\label{fig:unnamed-chunk-2}
\end{figure}

@@

We next tested whether NO signalling regulates UV/violet avoidance. To
achieve this, we generated two \emph{Platynereis} \emph{NOS} knockout
lines using the CRISPR/Cas9 method. We recovered two deletions
(NOSΔ11/Δ11 and Δ23/Δ23), both frame-shift mutations leading to an early
stop codon and thus likely representing null alleles (Figure 3---figure
supplement 1A). We could establish a homozygous line for both mutations
indicating that NOS is not an essential gene in \emph{Platynereis}. To
quantify UV avoidance, we recorded the trajectories of freely swimming
wild type and mutant larvae in vertical columns, illuminated laterally
from two opposite sides with 395 nm UV light (Figure 3A and Figure
3---figure supplement 1B). As previously shown, wild-type larvae swim
downward following non-directional UV/violet light stimulation (Verasztó
et al., 2018). In contrast, both two- and three-day-old homozygous
\emph{NOS}-mutant larvae showed a strongly diminished UV-avoidance
response (Figure 3A, B and Figure 3---figure supplement 1B, C). This
phenotype is similar to the defective UV-avoidance of \emph{c-opsin1}
mutant larvae (Verasztó et al., 2018) and reveals a requirement for NOS
in UV-avoidance behaviour. Wild type but not mutant larvae also showed
an increase in swimming speed under UV light that may be due to downward
swimming trajectories (swimming in the direction of gravity) (Figure 3B
and Figure 3---figure supplement 1C). We also tested directional
phototaxis, by exposing larvae to 480 nm directional collimated light
from the top of the column. Three-day-old but not two-day-old
\emph{NOS}-mutant larvae also showed reduced phototactic behaviour,
suggesting a function for \emph{NOS} in the visual eyes that mediate
three-day-old phototaxis (Randel et al., 2014) (Figure 3D and Figure
3---figure supplement 1G).

To distinguish between an acute and developmental function of NOS in
light responses, we also tested larvae exposed to the NOS inhibitor
L-NAME. Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a
dose-dependent inhibition of UV avoidance. In contrast, phototaxis was
not affected (Figure 3C, E). Overall, our results indicate an acute
requirement for NOS signalling in UV-avoidance and a possible indirect,
developmental role in the visual system, reminiscent of the function of
NO signalling in \emph{Drosophila} eye development (Gibbs and Truman,
1998).

\begin{figure}
\includegraphics[width=33.33in]{figures/Fig3} \caption{**Figure 3. NOS is required for UV avoidance in *Platynereis* larvae.**  (**A**) Swimming trajectories of wild type (WT, n=32) and *NOS* mutant (*NOSΔ11/Δ11*, n=26 and *NOSΔ23/Δ23*, n=47) three-day-old larvae. All trajectories start at 0 x and y position and time 0 corresponding to 10 sec after the onset of 395 nm stimulation from the side. (**B**) Vertical position of batches of wild type and mutant three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. (**C**) Vertical position of batches of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. (**D**) Vertical displacement in 30 sec bins of wild type and mutant (*NOSΔ11/Δ11* and *NOSΔ23/Δ23*) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. (**E**) Vertical displacement in 30 sec bins of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. <br> Figure 3 -- source data 1-5. Source data for panels A-E.}\label{fig:unnamed-chunk-3}
\end{figure}

@@

To investigate how NO signalling alters the dynamics of the cPRC
circuit, we carried out Ca\textsuperscript{2+} imaging experiments. We
ubiquitously expressed the Ca\textsuperscript{2+} sensor GCaMP6s in
larvae and imaged Ca\textsuperscript{2+} signals during 405 nm
stimulation of the cPRCs. As we have shown previously, a 20 sec local
stimulation of cPRC cilia led to a transient increase in cPRC
Ca\textsuperscript{2+} levels, followed by a transient decrease
(Verasztó et al., 2018). After \textasciitilde20-sec,
Ca\textsuperscript{2+} levels in cPRCs were raising again, reaching
higher levels than at the start of the stimulus -- a response that may
involve cPRC depolarisation (Figure 4A, B). This activation phase occurs
after the 20 sec stimulation period and is likely due to a delayed
neuroendocrine feedback (Verasztó et al., 2018). To determine whether NO
mediates such a feedback, we repeated the experiment in
\emph{NOS}-mutant larvae. While we detected the initial activation phase
followed by inhibition, in homozygous \emph{NOS}-mutants for both CRISPR
alleles this was not followed by delayed activation. Instead,
Ca\textsuperscript{2+} levels dropped to a low steady-state level
(Figure 4A, B). We thus identified a requirement for NO signalling in
the late-phase activation of cPRCs.
